Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-02-07 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1985 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 833d641068b635820b6d55dbdff06b99d2fa9671

\SweaveOpts{eval = TRUE}
\section{Introduction}
In regression analysis various models have been developed to describe the relationship between a response variable and a set of covariates. Starting from linear models (LM) being extended to generalized linear models (GLM) the limitation to model only linear effects has lead to the development of generalized additive models (GAM) also capturing non-linear effects. In Figure~\ref{Devel_parmod} this development is illustrated for the case of one covariate. In the left picture the expected value of the response variable is modeled by a linear function in the covariate. In the middle a GAM is applied where a smooth function allows for more flexibility.

These models consider only the expectation of the underlying distribution, however, in recent years there has been a paradigm shift from regression models for the mean to distributional regression approaches for probabilistic forecasting. For example in the field of precipitation forecasting linear models such as ensemble model output statistics (EMOS, \citealt{gneiting2005calibrated}) usually acting on the assumption of a normally distributed response variable allow for both parameters, the mean and the variance, to depend linearly on covariates. Based on the resulting model the probability of precipitation can be estimated as well as the expected amount of precipitation. Therefore, these models provide common and widely used tools, also due to their intuitive structure and interpretability. However, the restriction to model only linear effects has risen the interest in applying more flexible models. Generalized additive models for location, scale and shape (GAMLSS, \citealt{stasinopoulos2005}) offer a framework in which each distribution parameter can be modeled separately by its own GAM as illustrated in the right plot of Figure~\ref{Devel_parmod}. These flexible parametric methods provide well fitting models when applied on precipitation data in a rather smooth and homogeneous setting, e.g. areas where no abrupt changes in meteorological/climatological events appear. However, for example in complex terrain in mountainous regions various factors can change rapidly and their effects are difficult to be represented by a smooth parametric model. In this situation a different approach which can deal with these abrupt changes would be to apply tree-structured models such as regression trees (\citealt{breiman1984classification})
%%%<TH> citation Loh does not fit, maybe omit citations anyway </TH>
%(\citealt{loh2011classification}) 
or random forests such as Breiman and Cutler's randomForest (\citealt{breiman2001random}). Moreover, these models are particularly advantageous in case of many covariates with unknown effects or interactions as the variable selection is done automatically within the model fitting process. Therefore, no additional effort or knowledge for the specification of the model is required in advance. However, even though their main advantage of being able to model non-additive effects and interactions without any necessary prespecification can play to their strength, regression trees are not an adequate choice if (additional) smooth effects appear as well. In this situation a random forest provides a good compromise as it combines a set of trees such that the steps imposed by the trees are smoothed out in case of smooth effects while abrupt changes are still captured by the tree structure of the model.

@@

%\SweaveOpts{eval = TRUE}
\subsection{Application for one station}
First of all we focus on one observation station, Axams in Tyrol. This station is selected due to its geographical closeness to Innsbruck, the capital of Tyrol. As for all other stations, daily observations of the month of July of 28 years are available from 1985 through the year 2012.

The results of fitting one distributional forest to the first 24 years of the data set and predicting the total amount of precipitation for one specific day of each of the four successive years have been shown in Figure~\ref{axams_onestation} as a motivational example for a probabilistic forecast. Staying with this fitted model we are now going to analyze variable importance of the distributional forest and compare its results to those of the GAM models and the EMOS model based on residual QQ-plots. In the appendix the corresponding probability integral transform (PIT) histograms (\citealt{gneiting2007}) can be found.
%%% <TH> wouldn't be a uniform QQ-plot be more appropriate? </TH>

@@

In Figure~\ref{pit_oosample} another impression of the quality of the predictions made by the tested methods is given by residual QQ-plots for out-of-sample predictions of each of the models. For these plots 100 randomized quantile residuals are simulated per observation. The closer the points are to the diagonal line the better the estimated distributional model fits the data. Overall, it can be stated that all four models provide well calibrated forecasts.

@@

\subsection*{Probability integral transform (PIT) histograms}
\begin{figure}[t!]
\begin{center}
%\begin{subfigure}{0.8\textwidth}
<<rain_axams_pit_oosample, fig=TRUE, echo=FALSE, width = 6.5>>=

>>>>

\SweaveOpts{eval = TRUE}
\section{Introduction}
In regression analysis various models have been developed to describe the relationship between a response variable and a set of covariates. Starting from linear models (LM) being extended to generalized linear models (GLM) the limitation to model only linear effects has led to the development of generalized additive models (GAM) also capturing non-linear effects. In Figure~\ref{Devel_parmod} this development is illustrated for the case of one covariate. In the left picture the expected value of the response variable is modeled by a linear function in the covariate. In the middle a GAM is applied where a smooth function allows for more flexibility.

These models consider only the expectation of the underlying distribution. However, in recent years there has been a paradigm shift from regression models for the mean to distributional regression approaches for probabilistic forecasting. For example in the field of precipitation forecasting linear models such as ensemble model output statistics (EMOS, \citealt{gneiting2005calibrated}) usually acting on the assumption of a normally distributed response variable allow for both parameters, the mean and the variance, to depend linearly on covariates. Based on the resulting model the probability of precipitation can be estimated as well as the expected amount of precipitation. Therefore, these models provide common and widely used tools, also due to their intuitive structure and interpretability. However, the restriction to model only linear effects has risen the interest in applying more flexible models. Generalized additive models for location, scale and shape (GAMLSS, \citealt{stasinopoulos2005}) offer a framework in which each distribution parameter can be modeled separately by its own GAM as illustrated in the right plot of Figure~\ref{Devel_parmod}. These flexible parametric methods provide well fitting models when applied on precipitation data in a rather smooth and homogeneous setting, e.g. areas where no abrupt changes in meteorological/climatological events appear. However, for example in complex terrain in mountainous regions various factors can change rapidly and their effects are difficult to be represented by a smooth parametric model. In this situation a different approach which can deal with these abrupt changes would be to apply tree-structured models such as regression trees (\citealt{breiman1984classification})
%%%<TH> citation Loh does not fit, maybe omit citations anyway </TH>
%(\citealt{loh2011classification}) 
or random forests such as Breiman and Cutler's randomForest (\citealt{breiman2001random}). Moreover, these models are particularly advantageous in case of many covariates with unknown effects or interactions as the variable selection is done automatically within the model fitting process. Therefore, no additional effort or knowledge for the specification of the model is required in advance. However, even though their main advantage of being able to model non-additive effects and interactions without any necessary prespecification can play to their strength, regression trees are not an adequate choice if (additional) smooth effects appear as well. In this situation a random forest provides a good compromise as it combines a set of trees such that the steps imposed by the trees are smoothed out in case of smooth effects while abrupt changes are still captured by the tree structure of the model.

@@

%\SweaveOpts{eval = TRUE}
\subsection{Application for one station}
\label{Application for one station}
First of all we focus on one observation station, Axams in Tyrol. This station is selected due to its geographical closeness to Innsbruck, the capital of Tyrol. As for all other stations, daily observations of the month of July of 28 years are available from 1985 through the year 2012.

The results of fitting one distributional forest to the first 24 years of the data set and predicting the total amount of precipitation for one specific day of each of the four successive years have been shown in Figure~\ref{axams_onestation} as a motivational example for a probabilistic forecast. Staying with this fitted model we are now going to analyze variable importance of the distributional forest and compare its results to those of the GAM models and the EMOS model based on residual QQ-plots.
%%% <TH> wouldn't be a uniform QQ-plot be more appropriate? </TH>

@@

In Figure~\ref{qqr_oosample} another impression of the quality of the predictions made by the tested methods is given by residual QQ-plots for out-of-sample predictions of each of the models. For these plots 100 randomized quantile residuals are simulated per observation. The closer the points are to the diagonal line the better the estimated distributional model fits the data. Overall, it can be stated that all four models provide well calibrated forecasts. In the appendix the corresponding probability integral transform (PIT) histograms (\citealt{gneiting2007}) can be found.

@@

\subsection*{Probability integral transform (PIT) histograms}
Probability integral transform (PIT) histograms \cite{gneiting2007} are a commonly used tool to assess the predictive power of distributional forecasting methods, particularly in the field of meteorological forecasting \\
Let $x$ be an observation with true distribution function $G$ and predicted distribution function $F$. Then the corresponding probability integral transform (PIT) value is $p = F(x)$. In the ideal case of a perfect prediction and $F$ being continuous $p$ follows a uniform distribution. This uniformity can be analysed applying a histogram. The closer the bars of a PIT histogram are to the value 1, the closer the distribution of $p$ is to a uniform distribution and therefore the better the predicted distribution models the true distribution.

The PIT histograms in Figure~\ref{pit_oosample} correspond to the forecasts presented in Section \ref{Application for one station}. 
Daily observations of total precipitation from station Axams are considered and predictions are made for the 24th of July 2009, 2010, 2011 and 2012, based on data from the previous 24 years.

\begin{figure}[t!]
\begin{center}
%\begin{subfigure}{0.8\textwidth}
<<rain_axams_pit_oosample, fig=TRUE, echo=FALSE, width = 6.5>>=
